---
title: |
  Patterns of gaming behavior among NEET and non-NEET youth: A descriptive analysis of multi-platform telemetry
author:
  - name: Michael Trotter
    affiliations:
      - ref: 1
  - name: Jo Iacovides
    affiliations:
      - ref: 2
  - name: Regan Mandryk
    affiliations:
      - ref: 3
  - name: Nick Ballou
    corresponding: true
    orcid: 0000-0003-4126-0696
    email: nick@nickballou.com
    affiliations:
      - ref: 4
    roles:
      - conceptualization
      - data curation
      - investigation
      - methodology
      - formal analysis
      - visualization
      - writing
      - editing
affiliations:
  - id: 1
    name: Halmstad University
    country: Sweden
  - id: 2
    name: University of Victoria
    country: Canada
  - id: 3
    name: University of York
    country: United Kingdom
  - id: 4
    name: Imperial College London
    country: United Kingdom
abstract: |
 TODO
bibliography: references.bib
keywords:
  - video games
  - demographics
  - behavioral data
  - descriptive
  - genres
editor_options:
  chunk_output_type: console
  markdown:
    wrap: none
knitr:
  opts_chunk:
    cache: true
    warning: false
    message: false
    fig-width: 6
    fig-asp: 0.618
    out-width: 80%
    fig-align: center
code-links: repo
format:
  html:
    csl: https://www.zotero.org/styles/apa
    theme:
      light: united
      dark: darkly
    toc: true
    toc-depth: 2
    toc-location: right
    number-sections: true
    fontsize: 13pt
    code-fold: true
    code-summary: "Show code"
    code-tools: true
    echo: true
    embed-resources: false
    page-layout: article
    grid:
      body-width: 900px
      margin-width: 250px
    other-links:
      - text: Data
        href: https://zenodo.org/records/17609160
        icon: link
  preprint-typst:
    output-file: index.pdf
    echo: false
    wordcount: true
    theme: jou
    citeproc: true
    mainfont: "Liberation Serif"
    sansfont: "Liberation Sans"
    include-before-body:
      - text: "#show <refs>: set par(hanging-indent: 1.4em)"
  docx:
    echo: false
  # ACM format in _quarto-acm.yml - render with: quarto render --profile acm --to acm-pdf
---

```{r}
#| label: libraries
#| code-summary: "Show code (Load required R packages)"
#| cache: false
library(tidyverse)
library(glue)
library(lubridate)
library(ggdist)
library(patchwork)
library(cowplot)
library(tinytable)
library(effectsize)
library(MatchIt)
```

```{r}
#| label: setup
#| code-summary: "Show code (Set random seed and global options)"
#| cache: false
set.seed(8675309)
options(scipen = 999, timeout = 600)

theme_set(theme_minimal())
theme_update(
  plot.background = element_rect(fill = "white", color = NA),
  panel.background = element_rect(fill = "white", color = NA),
  strip.background = element_rect(fill = "black"),
  strip.text = element_text(color = "white", size = 10, face = "bold"),
  axis.text.y = element_text(color = "black", size = 10),
  axis.text.x = element_text(color = "black", size = 10),
  panel.grid.minor = element_blank(),
  panel.border = element_rect(
    colour = "black",
    fill = NA,
    linewidth = 1
  )
)

# Load helper functions
source("R/helpers.R")
```

```{r}
#| label: load-data
#| code-summary: "Show code (Download and load raw data files)"
zenodo_url <- "https://zenodo.org/records/20119134/files/digital-wellbeing/open-play-v1.2.1.zip?download=1"
zip_path <- "data/open-play-v1.2.1.zip"

if (!file.exists(zip_path)) {
  dir.create(dirname(zip_path), showWarnings = FALSE, recursive = TRUE)
  options(timeout = max(600, getOption("timeout")))

  message("Downloading 194MB file from Zenodo (may take a few minutes)...")
  tryCatch(
    {
      download.file(zenodo_url, zip_path, mode = "wb", method = "libcurl")
      message("Download complete!")
    },
    error = function(e) {
      message("Download failed. Please download manually from:")
      message(zenodo_url)
      message(glue("And save to: {normalizePath(zip_path, mustWork = FALSE)}"))
      stop(e)
    }
  )
}

# Extract zip if not already extracted
extract_dir <- "data/open-play-v1.2.1"
if (!dir.exists(extract_dir)) {
  message("Extracting zip archive...")
  unzip(zip_path, exdir = "data")
  # Rename the extracted folder to a simpler name
  extracted_folder <- list.dirs("data", recursive = FALSE, full.names = TRUE)
  extracted_folder <- extracted_folder[grepl(
    "digital-wellbeing-open-play",
    extracted_folder
  )]
  if (length(extracted_folder) == 1 && extracted_folder != extract_dir) {
    file.rename(extracted_folder, extract_dir)
  }
}

# Read raw data files
intake <- read_csv(
  file.path(extract_dir, "data/clean/survey_intake.csv.gz"),
  guess_max = 10000
)

biweekly <- readr::read_csv(
  file.path(extract_dir, "data/clean/survey_biweekly.csv.gz"),
  guess_max = 10000
)

surveys <- read_csv(
  file.path(extract_dir, "data/clean/survey_daily.csv.gz")
) |>
  filter(pid %in% intake$pid)

biweekly <- read_csv(
  file.path(extract_dir, "data/clean/survey_biweekly.csv.gz"),
  guess_max = 10000
)

xbox <- read_csv(file.path(extract_dir, "data/clean/xbox.csv.gz"))
nintendo <- read_csv(file.path(extract_dir, "data/clean/nintendo.csv.gz"))
steam <- read_csv(
  file.path(extract_dir, "data/clean/steam.csv.gz"),
  guess_max = 10000
)
games <- read_csv(
  file.path(extract_dir, "data/clean/game_metadata.csv.gz"),
  guess_max = 10000
)

# Load helper functions from the dataset (includes get_dst_offset for timezone handling)
source(file.path(extract_dir, "R/helpers.R"))
```

```{r}
#| label: prepare-telemetry
#| code-summary: "Show code (Process telemetry into session, hourly, daily, and weekly aggregations)"
# Cache directory for processed telemetry (speeds up re-renders)
cache_dir <- "data/cache"
cache_file <- file.path(cache_dir, "telemetry_cache.rds")

if (file.exists(cache_file)) {
  # Load cached objects
  message("Loading cached telemetry data...")
  cache <- readRDS(cache_file)
  sessions_telemetry <- cache$sessions_telemetry
  hourly_telemetry <- cache$hourly_telemetry
  daily_telemetry <- cache$daily_telemetry
  weekly_telemetry <- cache$weekly_telemetry
  telemetry_spans <- cache$telemetry_spans
  rm(cache)
} else {
  message("Processing telemetry data (this may take a few minutes)...")

  # Map participants to their local timezone (for DST-aware conversion)
  tz_map <- intake |>
    mutate(
      pid = as.character(pid),
      country,
      local_timezone,
      .keep = "none"
    ) |>
    distinct(pid, .keep_all = TRUE)

  # ---------------------------------------------------------------------------
  # SESSION-LEVEL (Nintendo + Xbox only; Steam doesn't provide session data)
  # ---------------------------------------------------------------------------
  sessions_telemetry <- bind_rows(
    xbox |> mutate(platform = "Xbox"),
    nintendo |> mutate(platform = "Nintendo")
  ) |>
    left_join(tz_map, by = "pid") |>
    filter(!is.na(local_timezone)) |>
    mutate(
      offset_start = get_dst_offset(session_start, country, local_timezone),
      offset_end = get_dst_offset(session_end, country, local_timezone),
      start_local = session_start + offset_start,
      end_local = session_end + offset_end,
      duration_min = as.numeric(difftime(
        session_end,
        session_start,
        units = "mins"
      ))
    ) |>
    filter(
      !is.na(session_start),
      !is.na(session_end),
      session_end > session_start,
      duration_min >= 1,
      title_id != "Wallpaper Engine" # not a game, but a desktop customization tool
    )

  # ---------------------------------------------------------------------------
  # HOURLY (all platforms)
  # ---------------------------------------------------------------------------
  hourly_from_sessions <- sessions_telemetry |>
    filter(!is.na(start_local), !is.na(end_local)) |>
    mutate(
      h0_local = floor_date(start_local, "hour"),
      h1_local = floor_date(end_local - seconds(1), "hour"),
      n_hours = as.integer(difftime(h1_local, h0_local, units = "hours")) + 1
    ) |>
    filter(!is.na(n_hours), n_hours > 0) |>
    tidyr::uncount(n_hours, .remove = FALSE, .id = "k") |>
    mutate(
      hour_start_local = h0_local + hours(k - 1),
      minutes = pmax(
        0,
        as.numeric(difftime(
          pmin(end_local, hour_start_local + hours(1)),
          pmax(start_local, hour_start_local),
          units = "mins"
        ))
      ),
      hour_start_utc = with_tz(hour_start_local, tzone = "UTC")
    ) |>
    select(
      pid,
      platform,
      title_id,
      hour_start_local,
      hour_start_utc,
      minutes
    ) |>
    group_by(pid, platform, title_id, hour_start_local, hour_start_utc) |>
    summarise(minutes = sum(minutes, na.rm = TRUE), .groups = "drop")

  hourly_from_steam <- steam |>
    select(pid, title_id, datetime_hour_start, minutes) |>
    mutate(pid = as.character(pid)) |>
    left_join(tz_map, by = "pid") |>
    filter(!is.na(local_timezone)) |>
    mutate(
      platform = "Steam",
      hour_start_utc = datetime_hour_start,
      offset = get_dst_offset(datetime_hour_start, country, local_timezone),
      hour_start_local = datetime_hour_start + offset
    ) |>
    select(
      pid,
      platform,
      title_id,
      hour_start_local,
      hour_start_utc,
      minutes
    ) |>
    group_by(pid, platform, title_id, hour_start_local, hour_start_utc) |>
    summarise(minutes = sum(minutes, na.rm = TRUE), .groups = "drop")

  hourly_telemetry <- bind_rows(hourly_from_sessions, hourly_from_steam)

  # Restrict each participant's data to ±3 months (91 days) of their intake survey date
  intake_window <- intake |>
    select(pid, intake_date = date) |>
    mutate(pid = as.character(pid))

  hourly_telemetry <- hourly_telemetry |>
    left_join(intake_window, by = "pid") |>
    filter(
      abs(as.numeric(difftime(
        as.Date(hour_start_local),
        intake_date,
        units = "days"
      ))) <=
        91
    ) |>
    select(-intake_date)

  # ---------------------------------------------------------------------------
  # DAILY (aggregated from hourly)
  # ---------------------------------------------------------------------------
  daily_telemetry <- hourly_telemetry |>
    mutate(day_local = as.Date(hour_start_local)) |>
    group_by(pid, platform, day_local) |>
    summarise(minutes = sum(minutes, na.rm = TRUE), .groups = "drop")

  # ---------------------------------------------------------------------------
  # WEEKLY (aggregated from daily)
  # ---------------------------------------------------------------------------
  weekly_telemetry <- daily_telemetry |>
    mutate(week = floor_date(day_local, "week")) |>
    group_by(pid, platform, week) |>
    summarise(minutes = sum(minutes, na.rm = TRUE), .groups = "drop")

  # ---------------------------------------------------------------------------
  # TELEMETRY SPANS (date ranges per participant/platform)
  # ---------------------------------------------------------------------------
  telemetry_spans <- daily_telemetry |>
    group_by(pid, platform) |>
    summarise(
      telemetry_start = min(day_local, na.rm = TRUE),
      telemetry_end = max(day_local, na.rm = TRUE) + hours(1),
      week = floor_date(telemetry_end, "week"),
      n_weeks = as.integer(difftime(
        telemetry_end,
        telemetry_start,
        units = "weeks"
      )) +
        1,
      .groups = "drop"
    )

  # Save to cache
  dir.create(cache_dir, showWarnings = FALSE, recursive = TRUE)
  saveRDS(
    list(
      sessions_telemetry = sessions_telemetry,
      hourly_telemetry = hourly_telemetry,
      daily_telemetry = daily_telemetry,
      weekly_telemetry = weekly_telemetry,
      telemetry_spans = telemetry_spans
    ),
    cache_file
  )
  message("Telemetry cache saved to ", cache_file)
}
```

```{r}
#| label: prepare-neet
#| code-summary: "Show code (Prepare NEET variables for analysis)"

dat <- intake |>
  mutate(
    gender = case_when(
      gender %in% c("Man", "Male") ~ "Man",
      gender %in% c("Woman", "Female") ~ "Woman",
      gender == "Prefer not to say" ~ NA_character_,
      is.na(gender) ~ NA_character_,
      TRUE ~ "Non-binary"
    ),
    # Ethnicity: harmonize UK/US categories
    ethnicity = case_when(
      ethnicity %in% c("White", "White alone") ~ "White",
      ethnicity %in%
        c(
          "Black",
          "Black or African American alone",
          "Black, African, Caribbean or Black British"
        ) ~ "Black",
      ethnicity %in%
        c("Asian", "Asian alone", "Asian or Asian British") ~ "Asian",
      ethnicity %in%
        c(
          "Mixed",
          "Mixed or multiple ethnic groups",
          "Two or More Races"
        ) ~ "Multiple",
      ethnicity %in% c("Prefer not to say") ~ NA_character_,
      is.na(ethnicity) ~ NA_character_,
      TRUE ~ "Other"
    ),
    unspecified_unemployed = tolower(employment) %in%
      c(
        "unemployed but looking",
        "unemployed looking for work",
        "not currently employed",
        "working occasionally for a fast-food delivery service, but otherwise unemployed",
        "was working till start of april",
        "unemployment",
        "unable to work.",
        "unable to work",
        "no",
        "looking for work",
        "looking for employment",
        "looking for a job.",
        "looking for a job.",
        "still looking for part-time job",
        "seeking employment, but not currently employed",
        "don't work",
        "actively looking for work"
      ),

    full_time_caretaker = tolower(employment) %in%
      c(
        "unpaid carer",
        "i am not able tow ork as ihave autistic children i may ahve to go to school for any time so i am their fulltime care giver",
        "full-time unpaid carer",
        "full time unpaid carer",
        "full time carer for other household member",
        "full time carer for disabled partner",
        "full time carer",
        "family carer",
        "family caregiver",
        "carer for parent",
        "carer",
        "carer for my autistic child and have disability myself",
        "caregiver",
        "a homemaker or stay-at-home parent"
      ),

    disabled = tolower(employment) %in%
      c(
        "ssdi",
        "ssi",
        "registered disabled",
        "workers comp",
        "unemployed/sick/disabled",
        "unemployed disabled",
        "temporarily off work due to disability",
        "unable to work due to disability",
        "unable to work disabled",
        "unable to work due to ongoing condition",
        "i am unable to work due to ill health",
        "i am unable to work due to health issues",
        "long term sickness",
        "longterm sickness",
        "long term sick leave",
        "long term sick",
        "long term illness",
        "legally disabled/unable to work",
        "not working due to health reasons",
        "not currently working due to health issues",
        "non working due to disability",
        "disabled/unable to work.",
        "disabled/unable to work",
        "disabled/sick",
        "disabled/ unable to work",
        "disabled unemployed",
        "disabled unable to work",
        "disabled unable to work.",
        "disabled on benefits",
        "disabled and unable to work",
        "disabled and unemployed",
        "disabled and not currently employed",
        "disabled and chronic illness",
        "disabled",
        "disability",
        "disability benefits",
        "disable",
        "disabed",
        "chronically ill"
      ),
  ) |>
  mutate(
    neet = (unspecified_unemployed | full_time_caretaker | disabled) & age <= 25
  )
```

```{r}
#| label: prepare-matching
#| code-summary: "Show code (1:1 propensity score matching of NEET to non-NEET participants)"
#| output: false
#| include: false
#| 
# Match each NEET participant to one non-NEET participant with similar age,
# gender, ethnicity, and country using nearest-neighbor PSM (logistic propensity model).
match_input <- dat |>
  filter(
    qualified,
    !is.na(age),
    !is.na(gender),
    !is.na(ethnicity),
    !is.na(country)
  )

m_out <- matchit(
  neet ~ age + gender + ethnicity + country,
  data = match_input,
  method = "nearest",
  distance = "glm",
  ratio = 1,
  replace = FALSE
)

summary(m_out, un = FALSE)

matched_pids <- match.data(m_out)$pid
dat_matched <- dat |> filter(pid %in% matched_pids)
```


# Introduction

We compare gaming behaviour between young people who are NEET (not in education, employment, or training; age ≤ 25) and a 1:1 propensity-score-matched sample of non-NEET players matched on age, gender, ethnicity, and country, using platform telemetry (Steam, Xbox, Nintendo) restricted to the three months before and after each participant's intake survey.

## Top games

The table below ranks the ten most-played games (by total playtime) separately for matched NEET and non-NEET players, alongside the number and percentage of each group who played the game, minutes per player, and genre metadata from IGDB.

```{r}
#| label: top-games

neet_totals <- dat_matched |>
  group_by(neet) |>
  summarise(total_players = n_distinct(pid))

# Derive genre, subgenre, and multiplayer labels from game metadata.
# games$original_name covers all platforms: game names for Steam/Nintendo, UUIDs for Xbox.
# IGDB data takes priority; Xbox-specific columns fill in where IGDB is missing.
game_meta <- games |>
  distinct(original_name, .keep_all = TRUE) |>
  transmute(
    title_id = original_name,
    genre = coalesce(str_extract(genres, "^[^,]+"), xbox_primary_genre),
    subgenre = coalesce(
      str_extract(str_remove(genres, "^[^,]+,\\s*"), "^[^,]+"),
      xbox_subgenre
    ),
    multiplayer = game_modes
  )

# Use distinct(pid, neet) so each participant contributes one row to the join,
# avoiding inflation of play time by the number of biweekly survey responses.
top_games <- dat_matched |>
  distinct(pid, neet) |>
  left_join(hourly_telemetry, by = "pid") |>
  group_by(neet, title_id) |>
  summarise(
    total_minutes = sum(minutes, na.rm = TRUE),
    n_players = n_distinct(pid),
    min_per_player = total_minutes / n_players,
    .groups = "drop"
  ) |>
  left_join(neet_totals, by = "neet") |>
  mutate(pct_players = 100 * (n_players / total_players)) |>
  left_join(game_meta, by = "title_id")

top_10 <- function(data, neet_val) {
  data |>
    filter(neet == neet_val, !is.na(title_id)) |>
    slice_max(n_players, n = 10) |>
    arrange(desc(n_players)) |>
    mutate(rank = row_number())
}

bind_rows(top_10(top_games, TRUE), top_10(top_games, FALSE)) |>
  select(
    rank,
    title_id,
    n_players,
    pct_players,
    total_minutes,
    min_per_player,
    genre,
    subgenre,
    multiplayer
  ) |>
  setNames(c(
    "Rank",
    "Game",
    "Players",
    "% Players",
    "Minutes",
    "Min/Player",
    "Genre",
    "Subgenre",
    "Mode(s)"
  )) |>
  tt(
    digits = 1,
    caption = "Top 10 games by total play time — matched NEET and non-NEET players"
  ) |>
  group_tt(i = list("NEET" = 1, "Non-NEET" = 11))
```

## Time of day

The figure below shows each group's share of total playtime across hours of the day, indicating when NEET and non-NEET players tend to game.

```{r}
#| label: fig-time-of-day
#| fig-cap: "Distribution of play across hours of the day for matched NEET and non-NEET players. Lines show each group's share of total playtime per hourly bin (timestamps in local time)."
#| code-summary: "Show code (time-of-day figure)"

hourly_neet <- hourly_telemetry |>
  inner_join(dat_matched |> distinct(pid, neet), by = "pid") |>
  mutate(hour = hour(hour_start_local)) |>
  group_by(neet, hour) |>
  summarise(total_minutes = sum(minutes, na.rm = TRUE), .groups = "drop") |>
  group_by(neet) |>
  mutate(prop = total_minutes / sum(total_minutes)) |>
  ungroup() |>
  mutate(group = if_else(neet, "NEET", "Non-NEET"))

ggplot(hourly_neet, aes(x = hour, y = prop, colour = group)) +
  geom_line(linewidth = 0.9) +
  scale_x_continuous(
    breaks = seq(0, 21, by = 3),
    labels = c("12am", "3am", "6am", "9am", "12pm", "3pm", "6pm", "9pm")
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_colour_manual(
    values = c("NEET" = "#E41A1C", "Non-NEET" = "#377EB8")
  ) +
  labs(x = "Hour of day", y = "Share of total playtime", colour = NULL) +
  theme(legend.position = "bottom")
```

## Distributions

The raincloud plots below compare the matched groups on six variables: objective playtime from telemetry, self-reported play, gaming disorder symptoms (GDT, wave 1), life satisfaction, and the perceived value of gaming for work and daily routines.

```{r}
#| label: fig-distributions
#| fig-cap: "Raincloud plots comparing matched NEET and non-NEET players on objective playtime (telemetry), self-reported play, gaming disorder symptoms (GDT, wave 1), and life satisfaction (mean across waves). Dots show individual participants; boxes show IQR and median; densities show the distribution shape."
#| code-summary: "Show code (distribution raincloud plots)"
#| fig-width: 8
#| fig-asp: 0.75

# Total telemetry hours in the ±91-day window
telemetry_hrs <- hourly_telemetry |>
  inner_join(dat_matched |> distinct(pid, neet), by = "pid") |>
  group_by(pid, neet) |>
  summarise(value = sum(minutes, na.rm = TRUE) / 60, .groups = "drop") |>
  mutate(variable = "Telemetry playtime (hours)")

# Helper: biweekly data for matched participants with neet status attached.
# Using biweekly directly (rather than dat_matched) avoids column-name conflicts
# from the biweekly/intake join, which suffixes shared columns like life_sat to .x/.y.
bw_matched <- biweekly |>
  filter(pid %in% matched_pids) |>
  left_join(dat_matched |> distinct(pid, neet), by = "pid")

# Self-reported biweekly play: mean across waves, minutes → hours
self_rep <- bw_matched |>
  group_by(pid, neet) |>
  summarise(
    value = mean(self_reported_biweekly_play, na.rm = TRUE) / 60,
    .groups = "drop"
  ) |>
  mutate(variable = "Self-reported play (hrs / 2 wks)")

# GDT mean at wave 1: recode text responses to numeric 1–5 then average items
gdt_levels <- c(
  "Never" = 1,
  "Rarely" = 2,
  "Sometimes" = 3,
  "Often" = 4,
  "Very often" = 5
)

gdt <- bw_matched |>
  filter(derived_wave == 1) |>
  mutate(across(gdt_1:gdt_4, ~ gdt_levels[.])) |>
  mutate(value = rowMeans(pick(gdt_1:gdt_4), na.rm = TRUE)) |>
  select(pid, neet, value) |>
  mutate(variable = "GDT mean (wave 1, 1\u20135)")

# Life satisfaction: mean across all available waves
life_sat_dat <- bw_matched |>
  group_by(pid, neet) |>
  summarise(value = mean(life_sat, na.rm = TRUE), .groups = "drop") |>
  mutate(variable = "Life satisfaction (0\u201310)")

# Gaming value (work): mean across all waves (-3 to 3)
gv_work <- bw_matched |>
  group_by(pid, neet) |>
  summarise(value = mean(gaming_value_work, na.rm = TRUE), .groups = "drop") |>
  mutate(variable = "Gaming value: work (\u22123\u20133)")

# Gaming value (routines): mean across all waves (-3 to 3)
gv_routines <- bw_matched |>
  group_by(pid, neet) |>
  summarise(
    value = mean(gaming_value_routines, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(variable = "Gaming value: routines (\u22123\u20133)")

plot_data <- bind_rows(
  telemetry_hrs,
  self_rep,
  gdt,
  life_sat_dat,
  gv_work,
  gv_routines
) |>
  filter(!is.na(value)) |>
  mutate(
    group = if_else(neet, "NEET", "Non-NEET"),
    variable = factor(
      variable,
      levels = c(
        "Telemetry playtime (hours)",
        "Self-reported play (hrs / 2 wks)",
        "GDT mean (wave 1, 1\u20135)",
        "Life satisfaction (0\u201310)",
        "Gaming value: work (\u22123\u20133)",
        "Gaming value: routines (\u22123\u20133)"
      )
    )
  )

ggplot(plot_data, aes(x = value, y = group, fill = group, colour = group)) +
  # Dots drawn first so they sit underneath density and box
  stat_dots(
    side = "bottom",
    overflow = "compress",
    dotsize = 0.5,
    alpha = 0.5
  ) +
  # normalize = "panels" prevents a spike in one panel from flattening all others.
  # density_bounded(bandwidth = "nrd0") avoids Sheather-Jones failures on sparse
  # integer data (gaming value items).
  stat_halfeye(
    side = "top",
    density = density_bounded(bandwidth = "nrd0"),
    adjust = 2,
    normalize = "panels",
    width = 0.5,
    .width = 0,
    point_colour = NA,
    alpha = 0.7
  ) +
  # Boxplot drawn last so it always sits on top of dots and density
  geom_boxplot(
    width = 0.08,
    outlier.shape = NA,
    fill = "white",
    colour = "grey20",
    linewidth = 0.5
  ) +
  facet_wrap(~variable, scales = "free_x", nrow = 3) +
  scale_fill_manual(values = c("NEET" = "#E41A1C", "Non-NEET" = "#377EB8")) +
  scale_colour_manual(values = c("NEET" = "#E41A1C", "Non-NEET" = "#377EB8")) +
  labs(x = NULL, y = NULL) +
  guides(fill = "none", colour = "none") +
  theme(panel.spacing = unit(12, "pt"))
```

